Dependent Object Types

### 2012-08-29 doc/candidacy/report.tex
first pass on ml paper

Their module system addresses some shortcomings of the SML module
system. Though the SML module system is higher-order, because SML
provides functors and substructures, it still treats modules as
second-class. For example, it isn't possible to return a structure
from an if statement. In fact, allowing this might require obscuring
the visibility of the resulting structure to ensure soundness. Another
limitation of the SML module system is that ``type sharing'' is
limited to equality between type names, not general type
expressions. In effect, this means that the transparency / opaqueness
of a signature cannot be fully fine-tuned. Their proposal introduces
translucent signatures, which allows any type to be partially
determined by a type expression.

%Scala's abstract types have close
%resemblances to abstract types of signatures in the module systems of
%ML.

>>>>

Their module system addresses some shortcomings of the SML module
system. Though the SML module system is higher-order, because SML
provides functors and substructures, it still treats modules as
second-class. For example, it isn't possible to return a structure
from an if statement. In fact, allowing this might require obscuring
the visibility of the resulting structure to ensure soundness. Another
limitation of the SML module system is that ``type sharing'' is
limited to equality between type names, not general type
expressions. In effect, this means that the transparency / opaqueness
of a signature cannot be fully fine-tuned. Their proposal introduces
translucent signatures, which allows any type to be partially or fully
determined by a type expression.

Scala's abstract types have close resemblances to abstract types of
signatures in these ML-like module systems. % TODO: develop

Their calculus is grounded in type theory. It is based on Girard's
$F_\omega$~\cite{tapl}. They add {\it translucent sums} to model
modules, dependent functions to model functors, a notion of subtyping
to model module implementation-interface matching. The crucial
addition and novelty is that of a {\it translucent sum}. A translucent
sum has the form of a possibly empty sequence of bindings. Similarly,
a translucent sum type has the form of a possibly empty sequence of
declarations. They differ from traditional records, because they can
contain types and because later fields can depend on ealier
ones. Translucent sums are ordinary values. Since they model modules,
it means that modules are first-class values. There's a form of
dependent type, since translucent sums are expressions that may
contain types.

In terms of theoretical results, Lillibridge shows in his thesis that
a simplified kernel system based on this calculus is sound, even in
the presence of side effects~\cite{lillibridge:thesis}. In the
appendix of the paper, they show that subtyping, and hence type
checking, is undecidable.
